### 2020-07-16 paper.Rmd
work on paper

---
title             : "A bicriterion extension of k-means anticlustering to optimize between-group similarity"
shorttitle        : "K-means anticlustering extended"

@@

abstract: |
  Anticlustering is an automated method to divide a data set into groups
  in such a way that high between-group similarity and high within-group 
  heterogeneity is obtained. Anticlustering is accomplished by maximizing 
  instead of minimizing a clustering objective.

@@

keywords          : "K-Means, Anticlustering, Variance"
wordcount         : "X"

@@

```{r setup, include = FALSE}
library("papaja")
library("anticlust")
library("knitr")
options(digits = 2)
knitr::opts_chunk$set(message=FALSE, warning = FALSE, echo = FALSE) 
```

@@

A different approach to anticlustering directly works with feature based 
data and does not require a transformation into pairwise 
dissimilarities. Instead, the sum of the squared Euclidean distances 
between elements and the cluster they are assigned to is minimized 
[brusco2006branch], most prominently using the k-means heuristic 
[@jain2010; @steinley2006]. K-means clustering maximizes between-cluster 
separation and within-cluster homogeneity simultaneously 
[@aloise2009np]. Späth [-@spath1986] and Valev [-@valev1998; 
-@valev1983] independently recognized that by optimizing the reversed 
k-means objective that is, maximizing the sum of distances between 
elements and cluster centroids -- a data set is partitioned into similar 
groups, thereby reversing the logic of classical clustering. They 
independently from each other coined the term anticlustering for this 
purpose. K-means anticlustering maximizes the sum of squared Euclidean 
distances from each element to the center of the cluster to which it 
belongs. As @spath1986 noted, k-means anticlustering thereby directly 
minimizes differences with regard to the means of the numeric variables 
between groups. Hence, groups only become similar with regard to the 
mean of the distribution of the numeric attributes, which may perform 
poorly as a measure of overall similarity; the spread of the data is not 
targeted and may differ between sets. @papenberg2020 showed that 
maximizing the diversity (based on the pairwise Euclidean distances) is 
better suited to minimize difference with regard to both the mean and 
the variance of the data; it was shown that even an entirely random 
split usually resulted in more similar standard deviations as k-means 
solely focuses on the mean. When requiring that groups are overall 
similar to each other, this focus may be detrimental as similar means 
can be obtained even when the underlying distributions are very 
different [@anscombe1973].

@@

## A Motivating Example

set.seed(4318) # 4318

N <- 14
K <- 2
partitions <- generate_partitions(N, K)
features <- round(rnorm(N, 15, 3), 1)

## Create an objective function that takes the partition
## as first argument (then, we can use sapply to compute
## the objective for each partition)
diff_mean <- function(clusters, features) {
  abs(diff(tapply(features, clusters, mean)))
}

diff_vars <- function(clusters, features) {
  abs(diff(tapply(features, clusters, var)))
}

all_objectives_mean <- sapply(
  partitions,
  FUN = diff_mean,
  features = features
)

plot(df, pch = 4, col = "darkgrey", las = 1)
# Illustrate best partition wrt simililarity in means
points(
  all_objectives_mean[which.min(all_objectives_mean)],
  all_objectives_var[which.min(all_objectives_mean)],
  cex = 2, col = "#DF536B", pch = 16
)

points(
  all_objectives_mean[which.min(df$sum)],
  all_objectives_var[which.min(df$sum)],
  cex = 2, col = "#2297E6", pch = 16
)

points(
  all_objectives_mean[which.min(df$diff_var)],
  all_objectives_var[which.min(df$diff_var)],
  cex = 2, col = "#61D04F", pch = 16
)

An motivating example illustrates that aiming to optimize similarity of 
means of standard deviations in a data set is not only desirable, but 
also expected to be feasible. To this end, I created 14 data points from 
a normal distribution ($M = 15$, $SD = 3$, see Table 1). The data points 
in Figure 1 illustrate all `r length(partitions)`  ways to partition a 
data set of $N = 14$ into two parts. Red point: maximum similarity in 
means; green point = maximum similarity in variance. Blue point: Almost 
no difference to red point wrt similarity in means, but variance is much 
more similar. This blue partition is certainly preferable to the red one 
when maximal overall similarity is required; the red one just optimizes 
similarity in means (which is what maximizing the k-means objective 
does).

```{r datatables}

best_partition_mean <- partitions[[which.min(df$diff_mean)]]
best_partition_var  <- partitions[[which.min(df$diff_var)]]
best_partition_combined <- partitions[[which.min(df$sum)]]

data <- data.frame(
  x = features,
  best_mean = best_partition_mean,
  best_var = best_partition_var,
  best_combined = best_partition_combined

)

descriptives <- cbind(
  mean_sd_tab(as.matrix(features), best_partition_mean),
  mean_sd_tab(as.matrix(features), best_partition_var),
  mean_sd_tab(as.matrix(features), best_partition_combined)
)

descriptives <- paste("M =", gsub("[(]", "(SD = ", descriptives))
dim(descriptives) <- c(2, 3)
colnames(descriptives) <- colnames(data[,-1])
apa_table(data, caption = "Illustrative data set")
apa_table(descriptives)

Unfortunately, improving algorithmic procedures will not help to find 
better partitions -- the objective function has to change. I will show 
an objective function that captures both similarity in means and 
variance.

## K-Means objective

In this treatment of k-means, I adopt the notation Steinley (2006) 
provided in his synthesis of the k-means literature. In clustering---or 
anticlustering---, an item pool $X$ = $\{x_1, \ldots, x_N\}$ has to be 
partitioned into $K$ groups $\{C_1, \ldots, C_K\}$. We assume that each 
element $x_i$ ($i = 1, \ldots, N$), is a vector of length $P$, and each 
entry describes one of its numeric measurements. Generally, $X$ 
can is interpreted as an $N \times M$ matrix where each row represents 
an data point $x_i$ (e.g., representing a stimulus or a person) and each 
column is a numeric attribute (e.g., the number of syllables in a word, 
or the attractiveness rating of a face).

K-Means anticlustering aims to maximize the error sum of squares (SSE), 
the within-group variance, which is the sum of the squared Euclidean 
distances between each data point and its cluster centroid. The cluster 
centroid is given by averaging the values on each variable over the 
elements within that cluster. The centroid of the *j*th variable in the 
*k*th cluster, consisting of $n_k$ elements, is given as:

$$
\overline{x}_j ^{(k)} = \frac{1}{n_k} \sum\limits {i \in C_k} x_{ij}
$$

The k-means objective function, the error sum of squares, is then given 
as:

@@

In anticlustering applications, we usually assume that all cluster are 
of equal size. Hence

## K-variances

@@

Important to note: We solve the same optimization problem, we just 
change the input a bit (we double the number of "columns").

**Example table!!**

@@

- formalize the weighted-sum objective and use the weighted-sum method.
works well on this specific problem [although in general there are 
caveats to the weighted-sum method, see @brusco2019]. is fast, good 
results. in fact, the objectives seem to work well together (optimizing 
k-means extended often yields more similar variances than just 
k-variances!!!)

@@

@@

- Standardization of the extended data matrix to ensure that the two
criterion functions do no differ markedly in scale [@brusco2019]

@@

>>>>

---
title             : "A bicriterion extension of k-means anticlustering to optimize between-group similarity"
shorttitle        : "Bicriterion K-means anticlustering"

@@

abstract: |
    Anticlustering is an automated method to partition data sets into groups 
    in such a way that high between-group similarity and high within-group 
    heterogeneity is obtained. Anticlustering has important applications in 
    psychological research, for example to partition stimulus sets into 
    equivalent sets, or to divide a group of students into heterogeneous 
    classes. K-means anticlustering is maximizes the error sum of squares, 
    i.e., the sum of squared Euclidean distances from each element to the 
    center of the group to which it belongs. Thereby, k-means anticlustering 
    directly maximizes the similarity of the mean values of numeric 
    attributes across groups, but ignores the spread of the data. 
    This paper presents a straight-forward extension of 
    the k-means objective function that allows to simultaneously maximize 
    similarity with regard to the mean as well as the variance across 
    groups. It is shown that when both criteria receive equal weights, the 
    bicriterion optimization problem reduces to an augmentation of the input 
    data, leaving the original k-means criterion unchanged. Tests on 
    simulated and real data show that similarity with regard to means and 
    variance criteria can usually be achieved to an astonishing degree with 
    almost no cost to either criterion. It is therefore concluded that the 
    extended model should be preferred to classical k-means anticlustering 
    for optimizing between-group similarity. An exception is given for 
    binary data that does not profit from considering equality in 
    variances.

@@

keywords          : "Anticlustering, variance, k-means, k-variance, bicriterion optimization"
wordcount         : "X"

@@

```{r setup, include = FALSE}
library("papaja")
library("anticlust")
library("knitr")
options(digits = 2)
knitr::opts_chunk$set(message=FALSE, warning = FALSE, echo = FALSE)

source("misc-functions.R")

@@

A different approach to anticlustering directly works with feature based 
data and does not require a transformation into pairwise 
dissimilarities. Instead, the sum of the squared Euclidean distances 
between elements and the cluster they are assigned to is minimized 
[@brusco2006branch], most prominently using the k-means heuristic 
[@jain2010; @steinley2006]. Minimizing the k-means criterion 
simultaneously maximizes between-cluster separation and within-cluster 
homogeneity [@aloise2009np]. Späth [-@spath1986] and Valev [-@valev1998; 
-@valev1983] independently recognized that by optimizing the reversed 
k-means objective that is, maximizing the sum of distances between 
elements and cluster centroids -- a data set is partitioned into similar 
groups, thereby reversing the logic of classical clustering. They 
independently from each other coined the term anticlustering for this 
purpose. K-means anticlustering maximizes the sum of squared Euclidean 
distances from each element to the center of the cluster to which it 
belongs. As @spath1986 noted, k-means anticlustering thereby directly 
minimizes differences with regard to the means of the numeric variables 
between groups. Hence, groups only become similar with regard to the 
mean of the distribution of the numeric attributes, which may perform 
poorly as a measure of overall similarity; the spread of the data is not 
targeted and may differ between sets. @papenberg2020 showed that 
maximizing the diversity (based on the pairwise Euclidean distances) is 
better suited to minimize difference with regard to both the mean and 
the variance of the data; it was shown that even an entirely random 
split usually resulted in more similar standard deviations as k-means 
solely focuses on the mean. When requiring that groups are overall 
similar to each other, this focus may be detrimental as similar means 
can be obtained even when the underlying distributions are very 
different [@anscombe1973].

@@

# Problem formalization

For the purpose of problem formalization, I adopt the notation Steinley 
(2006) provided in his thorough synthesis of a half-century literature 
on k-means. K-means clustering is used to partition two-way, two-mode 
data (that is, $N$ elements each having $P$ attributes) into $K$ groups 
$(C_1., \ldots, C_K)$ where $C_k$ is the set of $n_k$ objects in group 
$k$. The partitioning is subject to the requirements that each element 
is assigned to exactly one group, formalized by the two constraints

$$
\bigcup\limits_{k = 1}^{K} C_j = X
$$
$$
C_j \cap C_k = \emptyset, \; \forall j, k \in \{1, \ldots, K\}, \; j \ne k
$$

In anticlustering applications, we usually impose restrictions on 
the group sizes $n_k$, with the most common restriction being that all 
groups have equal size:

In this paper, I will discuss only such applications where groups are of 
equal size.

For a data matrix $\mathbf{X}_{N \times P} = \{x_{ij}\}_{N \times P}$, 
k-means clustering aims to minimize the error sum of squares (SSE)---the 
within-group variance---which is the sum of the squared Euclidean 
distances between each data point and its cluster centroid. Conversely, 
k-means anticlustering aims to maximize the same objective. The cluster 
centroid is given by averaging the values on each variable over the 
elements within that cluster. The centroid of the *j*th variable in the 
*k*th cluster is given as:

$$
\overline{x}_j ^{(k)} = \frac{1}{n_k} \sum\limits_{i \in C_k} x_{ij}
$$

The k-means objective function is then given as

@@

Note there is a direct connection between $SSE$ and the cluster 
centroids [@spath1986]. If the complete centroid vector for cluster 
$C_k$ is given by

$$
\overline{\mathbf{x}}^{(k)} = 
  (\overline{x}_1^{(k)}, \overline{x}_2^{(k)} \ldots, \overline{x}_P^{(k)})
$$

and the $SSE$ is minimal, the centroid values 
$\overline{\mathbf{x}}^{(k)} (k = 1, \ldots, K)$ are as far away as 
possible from the overall data centroid $\overline{\mathbf{x}} = 
(\overline{\mathbf{x}}_1, \overline{\mathbf{x}}_2, \ldots, 
\overline{\mathbf{x}}_P$), where $\overline{\mathbf{x}}_p$ is the 
overall mean value on the $p$th attribute:

$$
\overline{\mathbf{x}}_p = \frac{1}{N} \sum\limits_{i=1}^{N} x_{ip}
$$

If the $SSE$ is maximal---which is the goal of k-means 
anticlustering---the cluster centroids are as close as possible to the 
overall centroid $\overline{\mathbf{x}}$ and therefore to each other 
[@spath1986]. Thus, k-means anticlustering directly optimizes the 
similarity of the mean values of the $P$ attributes across the $K$ 
clusters.

## K-Variances

@@

## Criterion Tradeoff

# Generate data and all partitions

set.seed(2207) # 2207

N <- 14
K <- 2
partitions <- generate_partitions(N, K)
features <- rnorm(N, 15, 3)

# Compute objectives for each partition
all_objectives_mean <- sapply(
  partitions,
  FUN = diff_mean,
  features = features
)

A motivating example illustrates that aiming to simultaneously optimize 
similarity of means and variances at the same time is not only 
desirable, but may also be feasible. I created 14 data points from a 
univariate normal distribution ($M = 15$, $SD = 3$) and generated all `r 
length(partitions)` possible ways to partition the 14 data points into 
$K = 2$ equal-sized groups. I computed the difference in the two means 
and two variances for each partition. This simple difference metric was 
chosen for purposes of interpretability; in fact, there is a a 1-to-1 
correspondence between these differences and $SSE$ and 
$SSE_{\mathrm{VAR}}$, respectively (Figure 1). Figure 2 illustrates 
similarity in means plotted against similarity in variances across all 
partitions. Ideally, a partition is chosen that minimizes the 
difference with regard to means and variances (see the triangular 
symbol). However, optimal k-means anticlustering would create two groups 
that are not particularly similar with regard to their variance and 
chose the partition that is represented by the squared symbol. The 
diamond symbol represents that partition that where the two groups have 
the least difference in variances. The black symbols illustrate the 
other members of the optimal pareto set.

```{r, fig.cap = "Illustrates the 1-to-1 relationship between SSE and difference in means (Panel A) and variance (Panel B). Each data point illustrates a partition, i.e., a way to divide a data set ($N = 14$) into two groups.", fig.height = 3.5}

# For each partition: Compute SSE objective, and SSE(VAR)
# SSE
all_objectives_kMeans <- sapply(
  partitions,
  FUN = var_objective,
  features = features
)

# SSE(VAR)
all_objectives_kVar <- sapply(
  partitions,
  FUN = var_objective,
  features = squared_from_mean(as.matrix(features))
)

# Plot relationship between objectives
par(mfrow = c(1, 2))
plot(
  all_objectives_mean, 
  all_objectives_kMeans, 
  xlab = "mean(f1) - mean(f2)",
  ylab = "SSE",
  pch = 4, 
  cex = .5,
  col = "darkgrey"
)
legend("topright", legend = "A", cex = 1.3, bty = "n")
plot(
  all_objectives_var, 
  all_objectives_kVar, 
  xlab = "var(f1) - var(f2)",
  ylab = "SSE_var",
  pch = 4, 
  cex = .5,
  col = "darkgrey"
)
legend("topright", legend = "B", cex = 1.3, bty = "n")
```

```{r, fig.width = 4.5, fig.height = 4.5, fig.cap = "All partitions. Points = pareto efficient set."}

# Plot SSE vs SSE(VAR)
plot(
  df, pch = 4, col = "darkgrey", 
  las = 1, cex = .5,
  xlab = "mean(f1) - mean(f2)",
  ylab = "var(f1) - var(f2)"
)

# Illustrate Pareto set
pareto_pts <- pareto_set(df, sense = "min")
points(df[pareto_pts, ], pch = 19)

# Illustrate best partition wrt SSE (difference in means)
points(
  all_objectives_mean[which.min(all_objectives_mean)],
  all_objectives_var[which.min(all_objectives_mean)],
  cex = 1.2, pch = 22, bg = "white", lwd = 1.7
)

# Illustrate best partition wrt SSE(VAR) (i.e., difference in variance)
points(
  all_objectives_mean[which.min(df$sum)],
  all_objectives_var[which.min(df$sum)],
  cex = 1.2, pch = 24, bg = "white", lwd = 1.7
)

# Illustrate best partition wrt simililarity combined similarity / variance
points(
  all_objectives_mean[which.min(df$diff_var)],
  all_objectives_var[which.min(df$diff_var)],
  cex = 1.2, pch = 23, bg = "white", lwd = 1.7
)
```

@@

```{r echo = FALSE}
# From Wikipedia on Multi-objective optimization:

# Without additional subjective preference information, all Pareto optimal 
# solutions are considered equally good. Researchers study multi-objective 
# optimization problems from different viewpoints and, thus, there exist 
# different solution philosophies and goals when setting and solving them. 
# The goal may be to find a representative set of Pareto optimal 
# solutions, and/or quantify the trade-offs in satisfying the different 
# objectives, and/or finding a single solution that satisfies the 
# subjective preferences of a human decision maker (DM).

# The most preferred results can be found using different philosophies. 
# Multi-objective optimization methods can be divided into four 
# classes. In so-called no preference methods, no DM is expected to be 
# available, but a neutral compromise solution is identified without 
# preference information. The other classes are so-called a priori, a 
# posteriori and interactive methods and they all involve preference 
# information from the DM in different ways.

# Solving a multi-objective optimization problem is sometimes understood 
# as approximating or computing all or a representative set of Pareto 
# optimal solutions.

# Scalarizing a multi-objective optimization problem is an a priori 
# method, which means formulating a single-objective optimization problem 
# such that optimal solutions to the single-objective optimization problem 
# are Pareto optimal solutions to the multi-objective optimization 
# problem. In addition, it is often required that every Pareto optimal 
# solution can be reached with some parameters of the scalarization.

- pragmatic solution [easy and practical approach, @marler2010; 
@naidu2014]: formalize the weighted-sum objective and use the 
weighted-sum approach. Then check out whether the simplest version is 
sufficient in this case (but see the problems of weighted sum approach 
in several contributions by Brusco et al.). Refer to @marler2010 to 
argue for weighted sum for the current problem

- it should be noted that both mathematical objectives capture different 
aspects of an underlying substantive objective, that is to maximize 
similarity of data distributions between different groups

- Weighted sum approach works well on this specific problem [although in 
general there are caveats to the weighted-sum method, see @brusco2019]. 
is fast, good results. in fact, the objectives seem to work well 
together

- weighted sum approach generates a single solution from the 
(approximated) pareto set

@@

### Weight selection

@@

- Preference to weight [e.g., @marler2010]. However, the relationship 
between weights and outcome is not always straightforward and the 
results may disappoint. On the one hand, vastly different weights need 
not lead to different solutions; and even equal weights can be 
sufficient if the solution space offers partitions that maximize both 
objectives at the same time. "the solution may not preserve one’s
initial preferences no matter how the weights are set"; "even if one 
determines acceptable values for the weights a priori, the final 
solution may not accurately reflect initial preferences" [@marler2010, 
p. 860].

- Standardization of the extended data matrix to ensure that the two
criterion functions "do not differ markedly in scale" [@brusco2019]. 
See Figure 1: the objectives do differ in scale, because k-variance 
is based on a quadratic distance and therefore has larger values.

@@

# Discussion

- Focus on presentation of new objective function, the bicriterion 
optimization process might be improved [e.g., by using a direct 
algorithm as @brusco2019 did]

- new criterion: optimizes between-group similarity, no correspondence 
to within-group heterogeneity [see @brusco2019 for a exceptional method 
maximizing within-group heterogeneity]

- Similarly: the reversed objective no longer has a useful clustering 
interpretation: Maximizing differences in variances does not lead to 
homogeneous and/or well separated clusters
